
# 📊 Descriptive Statistics

```{r, echo=FALSE, include=FALSE}
library(here)
library(ayuda)
library(ggplot2)
source(here::here("R/init.R"))
```

```{r, echo=FALSE}
#Dataset prep
adlb <- adam$adlb() %>% 
  filter(SAFFL == "Y" &  AVISITN==9999 & ANL01FL == "Y" & !(PARAMCD %in% c("COLOR"))) %>%  # nolint
  select(USUBJID, AGE, TRT01P, PARCAT1, PARAM, PARAMCD, PARAMN, AVISIT, AVISITN, AVAL, CHG, BASE) # nolint

UNIQ_PARAMs <- adlb  %>% select(PARCAT1, PARAM, PARAMCD, PARAMN)  %>% unique()

## List unique Parameters
adlb_wide <- adlb %>%
  select(USUBJID, AGE, TRT01P, PARAMCD, CHG, BASE, AVAL, AVISIT) %>%
  pivot_wider(
    id_cols = c(USUBJID, AGE, TRT01P, AVISIT),
    names_from = PARAMCD,
    values_from = c(AVAL, CHG, BASE)
  )
```

## Sample Size
- **Definition:** Number of observations in the dataset  
- **Formula:**
$$
n
$$

<style>
.container {
  display: flex;
  gap: 20px; /* increase spacing */
}
.box {
  flex: 1;
  padding: 15px;
  border: 1px solid #ddd;
  border-radius: 6px;
}
</style>

<div class="container">

<div class="box">

##### Variable BASE_CA
```{r}
get_obs(adlb_wide$BASE_CA) %>% df2print()
```
</div> 

<div class="box">

##### Variable AGE
```{r}
get_obs(adlb_wide$AGE) %>% df2print()
```
</div> </div>

---

## Histogram and Range
- **Range:** Difference between max and min  
- **Formula:**
$$
\text{Range} = x_{\max} - x_{\min}
$$

#### Constructing a Histogram (Step-by-Step)

- Find the minimum value $x_{\min}$
- Find the maximum value $x_{\max}$
- Choose Number of Classes:
  - Classes are the groups (bins) used in a histogram.
  
<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
<strong>Rule of Thumb (Sturges’ Rule):</strong><br>
$$k = 1 + 3.322 \log_{10}(n)$$
- Where:
    - \(k\) = number of classes  
    - \(n\) = number of observations  

    Simple guideline:

    | Sample Size (n) | Number of Classes (k) |
    |----------------|------------------------|
    | 10–20          | 4–5                    |
    | 20–50          | 5–7                    |
    | 50–100         | 7–10                   |
    | 100+           | 10–20                  |
</div>
 

- Compute Class Width:
  - Formula:
    \[
    \text{Class Width} = \frac{\text{Range}}{\text{Number of Classes}}
    \]

- Create Class Intervals:
  - No overlapping intervals
  - Cover all data values

- Calculate Relative Frequency:
  - Formula:
  \[
  \text{Relative Frequency} = \frac{f}{n}
  \]

- Draw the Histogram

<div class="container">

  <div class="box">

##### Variable BASE_CA
  ```{r}
  get_range(adlb_wide$BASE_CA) %>% df2print()
  hist(adlb_wide$BASE_CA, col = "lightblue")
  ```
  </div> 

  <div class="box">

##### Variable AGE
  ```{r}
  get_range(adlb_wide$AGE) %>% df2print()
  hist(adlb_wide$AGE, col = "lightblue")
  ```
  </div>
</div>

---

## Central Tendency (average, median, mode)
#### - Average (mean)

  - **Definition:** Sum of all values divided by number of observations  
  - **Formula:**
$$
\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i
$$

#### - Median
  - **Definition:** Middle value when data is sorted  

- If \( n \) is odd:
$$
\text{Median} = x_{\frac{n+1}{2}}
$$

- If \( n \) is even:
$$
\text{Median} = \frac{x_{\frac{n}{2}} + x_{\frac{n}{2}+1}}{2}
$$

#### - Mode
  - **Definition:** Most frequently occurring value
<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
The mode is preffered over the mean and median only if the relative frequency 
of occurence of $y$ is of interest. For example, a supplier of carpenters materials would
be intersted in the modal lenght of neils he sells.
</div>

<div class="container">

  <div class="box">

##### Variable BASE_CA
  ```{r}
  get_central(adlb_wide$BASE_CA) %>% df2print() 
  ```
  </div> 

  <div class="box">

##### Variable AGE
  ```{r}
  get_central(adlb_wide$AGE) %>% df2print()
  ```
  </div>
</div>

---

## Variance & Standard Deviation

#### - Population Variance 
$$
\sigma^2 = \frac{1}{N} \sum_{i=1}^{N} (x_i - \mu)^2
$$

#### - Population Standard Deviation
$$
\sigma = \sqrt{\frac{1}{N} \sum (x_i - \mu)^2}
$$

#### - Sample Variance
$$
s^2 = \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2
$$
See details in Appendix \@ref(appendix-a).

#### - Sample Standard Deviation
$$
s = \sqrt{s^2}
$$

<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
  <strong>The Emperiical Rule:</strong><br>

  - Approximately 68% of data falls within 1 standard deviation of the mean $(\bar{x} \pm s)$ 
  - Approximately 95% of data falls within 2 standard deviations of the mean $(\bar{x} \pm 2s)$
  - Almost all data (99.7%) falls within 3 standard deviations of the mean $(\bar{x} \pm 3s)$.
</div>

<div class="container">

  <div class="box">

##### Variable BASE_CA
  ```{r}
  get_sd(adlb_wide$BASE_CA) %>% df2print() 
  ```
  </div> 

  <div class="box">

##### Variable AGE
  ```{r}
  get_sd(adlb_wide$AGE) %>% df2print()
  ```
  </div>
</div>

---

## Coefficient of Variation
- **Definition:** Relative measure of variability  
- **Formula:**
$$
CV = \frac{s}{\bar{x}} \times 100\%
$$

It measures the $std$ relative to the mean.

<div class="container">

  <div class="box">

##### Variable BASE_CA
  ```{r}
  100*sd(adlb_wide$BASE_CA)/mean(adlb_wide$BASE_CA) 
  ```
  </div> 

  <div class="box">

##### Variable AGE
  ```{r}
  100*sd(adlb_wide$AGE)/mean(adlb_wide$AGE) 
  ```
  </div>
</div>

---

## Quartiles
- **Definition:** Divide data into 4 equal parts  

- Q1: 25th percentile  
- Q2: Median (50th percentile)  
- Q3: 75th percentile  

<div class="container">

  <div class="box">

##### Variable BASE_CA
  ```{r}
  get_quartiles(adlb_wide$BASE_CA) %>% df2print() 
  ```
  </div> 

  <div class="box">

##### Variable AGE
  ```{r}
  get_quartiles(adlb_wide$AGE) %>% df2print()
  ```
  </div>
</div>

---

## Interquartile Range (IQR)
- **Definition:** Distance between Q1 and Q3  
- **Formula:**
$$
IQR = Q3 - Q1
$$

#### - Inner and outer fences
$$
\text{Inner fences} = [Q1 - 1.5 \times IQR, Q3 + 1.5 \times IQR]
$$
$$
\text{Outer fences} = [Q1 - 3 \times IQR, Q3 + 3 \times IQR]
$$

<div class="container">

  <div class="box">

##### Variable BASE_CA
  ```{r}
  get_iqr(adlb_wide$BASE_CA) %>% df2print() 
  boxplot(adlb_wide$BASE_CA, col = "lightblue", horizontal = TRUE)
  ```
  </div> 

  <div class="box">

##### Variable AGE
  ```{r}
  get_iqr(adlb_wide$AGE) %>% df2print()
  boxplot(adlb_wide$AGE, col = "lightblue", horizontal = TRUE)
  ```
  </div>
</div>

---

## Z-score (Standard Score)
- **Definition:** A z-score shows how many standard deviations a value is from the mean. 
- **Formula:**
$$
z = \frac{x - \bar{x}}{s}
$$


<div class="container">

  <div class="box">

##### Variable BASE_CA
  ```{r} 
  z <- scale(adlb_wide$BASE_CA)
  hist(z, main = "Distribution of Z-scores",
       xlab = "Z-score",
       col = "lightblue",
       breaks = 5)
  ```
  </div> 

  <div class="box">

##### Variable AGE
  ```{r} 
  z <- scale(adlb_wide$AGE)
  hist(z, main = "Distribution of Z-scores",
       xlab = "Z-score",
       col = "lightblue",
       breaks = 5)
  ```
  </div>
</div>

<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
  <strong>Rule of thumb for detecting **outliers**:</strong><br>

  - **z-score**: observations with |z| > 3 are considered **outliers**;
  - **Box plot**: observations falling between the inner and outer fences are considered **mild outliers**; observations falling outside the outer fences are considered **extreme outliers**.
</div>

## 11. Skewness
- **Definition:** Measure of asymmetry  
- **Formula:**
$$
\text{Skewness} = \frac{1}{n} \sum \left(\frac{x_i - \bar{x}}{s}\right)^3
$$

---

## 12. Kurtosis
- **Definition:** Measure of tail heaviness  
- **Formula:**
$$
\text{Kurtosis} = \frac{1}{n} \sum \left(\frac{x_i - \bar{x}}{s}\right)^4
$$

---

## 13. Covariance
- **Definition:** Joint variability of two variables  
- **Formula:**
$$
\text{Cov}(X,Y) = \frac{1}{n-1} \sum (x_i - \bar{x})(y_i - \bar{y})
$$

---

## 14. Correlation (Pearson)
- **Definition:** Standardized covariance  
- **Formula:**
$$
r = \frac{\text{Cov}(X,Y)}{s_x s_y}
$$

---
