
# 📊 Descriptive Statistics

```{r, echo=FALSE, include=FALSE}
library(here)
library(ayuda)
library(ggplot2)
source(here::here("R/init.R"))
```

```{r, echo=FALSE}
#Dataset prep
adlb <- adam$adlb() %>% 
  filter(SAFFL == "Y" &  AVISITN==9999 & ANL01FL == "Y" & !(PARAMCD %in% c("COLOR"))) %>%  # nolint
  select(USUBJID, AGE, TRT01P, PARCAT1, PARAM, PARAMCD, PARAMN, AVISIT, AVISITN, AVAL, CHG, BASE) # nolint

UNIQ_PARAMs <- adlb  %>% select(PARCAT1, PARAM, PARAMCD, PARAMN)  %>% unique()

## List unique Parameters
adlb_wide <- adlb %>%
  select(USUBJID, AGE, TRT01P, PARAMCD, CHG, BASE, AVAL, AVISIT) %>%
  pivot_wider(
    id_cols = c(USUBJID, AGE, TRT01P, AVISIT),
    names_from = PARAMCD,
    values_from = c(AVAL, CHG, BASE)
  )

A <- adlb_wide$BASE_CA
B <- adlb_wide$CHG_CA

```

## Sample Size
- **Definition:** Number of observations in the dataset  
- **Formula:**
$$
n
$$

<div class="container">

<div class="box">

**Variable A**
```{r}
get_obs(A) %>% df2print()
```
</div> 

<div class="box">

**Variable B**
```{r}
get_obs(B) %>% df2print()
```
</div> </div>

---

## Histogram and Range
- **Range:** Difference between max and min  
- **Formula:**
$$
\text{Range} = x_{\max} - x_{\min}
$$

#### Constructing a Histogram (Step-by-Step)

- Find the minimum value $x_{\min}$
- Find the maximum value $x_{\max}$
- Choose Number of Classes:
  - Classes are the groups (bins) used in a histogram.
  
<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
<strong>Rule of Thumb (Sturges’ Rule):</strong><br>
$$k = 1 + 3.322 \log_{10}(n)$$
- Where:
    - \(k\) = number of classes  
    - \(n\) = number of observations  

    Simple guideline:

    | Sample Size (n) | Number of Classes (k) |
    |----------------|------------------------|
    | 10–20          | 4–5                    |
    | 20–50          | 5–7                    |
    | 50–100         | 7–10                   |
    | 100+           | 10–20                  |
</div>
 


- Compute Class Width:
  - Formula:
    \[
    \text{Class Width} = \frac{\text{Range}}{\text{Number of Classes}}
    \]

- Create Class Intervals:
  - No overlapping intervals
  - Cover all data values

- Calculate Relative Frequency:
  - Formula:
  \[
  \text{Relative Frequency} = \frac{f}{n}
  \]

- Draw the Histogram

<div class="container">

  <div class="box">

**Variable A**
  ```{r}
  get_range(A) %>% df2print()
  hist(A, col = "lightblue")
  ```
  </div> 

  <div class="box">

**Variable B**
  ```{r}
  get_range(B) %>% df2print()
  hist(B, col = "lightblue")
  ```
  </div>
</div>

---

## Central Tendency (average, median, mode)
#### - Average (mean)

  - **Definition:** Sum of all values divided by number of observations  
  - **Formula:**
$$
\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i
$$

#### - Median
  - **Definition:** Middle value when data is sorted  

- If \( n \) is odd:
$$
\text{Median} = x_{\frac{n+1}{2}}
$$

- If \( n \) is even:
$$
\text{Median} = \frac{x_{\frac{n}{2}} + x_{\frac{n}{2}+1}}{2}
$$

#### - Mode
  - **Definition:** Most frequently occurring value
<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
The mode is preffered over the mean and median only if the relative frequency 
of occurence of $y$ is of interest. For example, a supplier of carpenters materials would
be intersted in the modal lenght of neils he sells.
</div>

<div class="container">

  <div class="box">

**Variable A**
  ```{r}
  get_central(A) %>% df2print() 
  ```
  </div> 

  <div class="box">

**Variable B**
  ```{r}
  get_central(B) %>% df2print()
  ```
  </div>
</div>

---

## Variance & Standard Deviation

#### - Population Variance 
$$
\sigma^2 = \frac{1}{N} \sum_{i=1}^{N} (x_i - \mu)^2
$$

#### - Population Standard Deviation
$$
\sigma = \sqrt{\frac{1}{N} \sum (x_i - \mu)^2}
$$

#### - Sample Variance
$$
s^2 = \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2
$$
See details in Appendix \@ref(appendix-a).

#### - Sample Standard Deviation
$$
s = \sqrt{s^2}
$$

<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
  <strong>The Emperiical Rule:</strong><br>

  - Approximately 68% of data falls within 1 standard deviation of the mean $(\bar{x} \pm s)$ 
  - Approximately 95% of data falls within 2 standard deviations of the mean $(\bar{x} \pm 2s)$
  - Almost all data (99.7%) falls within 3 standard deviations of the mean $(\bar{x} \pm 3s)$.
</div>


<div class="container">

  <div class="box">

**Variable A**
  ```{r}
  get_sd(A) %>% df2print() 
  ```
  </div> 

  <div class="box">

**Variable B**
  ```{r}
  get_sd(B) %>% df2print()
  ```
  </div>
</div>

---

## Coefficient of Variation
- **Definition:** Relative measure of variability  
- **Formula:**
$$
CV = \frac{s}{\bar{x}} \times 100\%
$$

It measures the $std$ relative to the mean.

<div class="container">

  <div class="box">

**Variable A**
  ```{r}
  100*sd(A)/mean(A) 
  ```
  </div> 

  <div class="box">

**Variable B**
  ```{r}
  100*sd(B)/mean(B) 
  ```
  </div>
</div>

---

## Quartiles
- **Definition:** Divide data into 4 equal parts  

- Q1: 25th percentile  
- Q2: Median (50th percentile)  
- Q3: 75th percentile  

<div class="container">

  <div class="box">

**Variable A**
  ```{r}
  get_quartiles(A) %>% df2print() 
  ```
  </div> 

  <div class="box">

**Variable B**
  ```{r}
  get_quartiles(B) %>% df2print()
  ```
  </div>
</div>

---

## Interquartile Range (IQR)
- **Definition:** Distance between Q1 and Q3  
- **Formula:**
$$
IQR = Q3 - Q1
$$

#### - Inner and outer fences
$$
\text{Inner fences} = [Q1 - 1.5 \times IQR, Q3 + 1.5 \times IQR]
$$
$$
\text{Outer fences} = [Q1 - 3 \times IQR, Q3 + 3 \times IQR]
$$

<div class="container">

  <div class="box">

**Variable A**
  ```{r}
  get_iqr(A) %>% df2print() 
  boxplot(A, col = "lightblue", horizontal = TRUE)
  ```
  </div> 

  <div class="box">

**Variable B**
  ```{r}
  get_iqr(B) %>% df2print()
  boxplot(B, col = "lightblue", horizontal = TRUE)
  ```
  </div>
</div>

---

## Z-score (Standard Score)
- **Definition:** A z-score shows how many standard deviations a value is from the mean. 
- **Formula:**
$$
z = \frac{x - \bar{x}}{s}
$$


<div class="container">

  <div class="box">

**Variable A**
  ```{r} 
  z <- scale(A)
  hist(z, main = "Distribution of Z-scores",
       xlab = "Z-score",
       col = "lightblue",
       breaks = 5)
  ```
  </div> 

  <div class="box">

**Variable B**
  ```{r} 
  z <- scale(B)
  hist(z, main = "Distribution of Z-scores",
       xlab = "Z-score",
       col = "lightblue",
       breaks = 5)
  ```
  </div>
</div>



<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
  <strong>Rule of thumb for detecting **outliers**:</strong><br>

  - **z-score**: observations with |z| > 3 are considered **outliers**;
  - **Box plot**: observations falling between the inner and outer fences are considered **mild outliers**; observations falling outside the outer fences are considered **extreme outliers**.
</div>


## Skewness
- **Definition:** Measure of asymmetry  
More probability than expected in one tail of the distribution.
- **Formula:**
$$
\text{Skewness} = \frac{1}{n} \sum \left(\frac{x_i - \bar{x}}{s}\right)^3
$$

| Skewness Value | Meaning            | Interpretation |
|----------------|-------------------|----------------|
| ≈ 0            | Symmetric         | Data is evenly distributed; mean ≈ median |
| > 0            | Positive skew     | Right-skewed; long tail on the right; mean > median |
| < 0            | Negative skew     | Left-skewed; long tail on the left; mean < median |
| 0.5 to 1       | Moderate skew     | Noticeable asymmetry in distribution |
| > 1            | Strong skew       | Highly asymmetric distribution |


<div class="container">

  <div class="box">

**Variable A**
  ```{r} 
  get_skewness(A, na.rm = TRUE) 
  ```
  </div> 

  <div class="box">

**Variable B**
  ```{r} 
  get_skewness(B, na.rm = TRUE) 
  ```
  </div>
</div>

---

## Kurtosis
- **Definition:** Measure of tail heaviness  
Probability are pushed away from the mean towards the tails.
- **Formula:**
$$
\text{Raw Kurtosis} = \frac{1}{n} \sum \left(\frac{x_i - \bar{x}}{s}\right)^4
$$
$$
\text{Excess Kurtosis} = \text{Raw Kurtosis} - 3
$$

| Kurtosis Value (Excess) | Distribution Type | Meaning | Interpretation |
|--------------------------|------------------|---------|----------------|
| ≈ 3 (Raw kurtosis)       | Mesokurtic          | Normal-like distribution | Similar peak and tail thickness as normal distribution |
| > 3 (Raw kurtosis)       | Leptokurtic         | Heavy-tailed distribution | More extreme values (outliers), sharper peak |
| < 3 (Raw kurtosis)       | Platykurtic         | Light-tailed distribution | Fewer outliers, flatter peak, more evenly spread data |
| ≈ 0 (Excess kurtosis)    | Mesokurtic       | Normal-like distribution | Similar peak and tails as normal distribution |
| > 0 (Excess kurtosis)    | Leptokurtic      | Heavy-tailed distribution | More outliers; sharper peak than normal |
| < 0 (Excess kurtosis)    | Platykurtic      | Light-tailed distribution | Fewer outliers; flatter peak than normal |

<div class="container">

  <div class="box">

**Variable A**
  ```{r} 
  get_kurtosis(A, na.rm = TRUE) 
  ```
  </div> 

  <div class="box">

**Variable B**
  ```{r} 
  get_kurtosis(B, na.rm = TRUE) 
  ```
  </div>
</div>

---

## Covariance
- **Definition:** Joint variability of two variables  
Covariance is one of family of statistical measures that describe the relationship between two variables. 
It indicates the direction of the linear relationship between them. 
A positive covariance indicates that the two variables tend to increase together, while a negative covariance indicates that one variable tends to increase when the other decreases. 
However, covariance does not provide information about the strength of the relationship, which is why correlation is often used as a standardized measure of covariance.
- **Formula:**
$$
\text{Cov}(X,Y) = \frac{1}{n-1} \sum (x_i - \bar{x})(y_i - \bar{y})
$$

<div class="container">

  <div class="box">

**Variable A vs B**
  ```{r} 
  cov(A, B)
  ```
  </div> 

</div>

Covariance result has no upper or lower bound, and its value is influenced by the scale of the variables. 
While correlation, on the other hand, is a standardized measure of covariance that ranges between -1 and 1, providing information about both the direction and 
strength of the linear relationship between two variables.

---

## Correlation (Pearson) {#correlation}
- **Definition:** Standardized covariance  

<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
  <strong>Correlation caveats:</strong><br>

  - Look at you data(scatterplot) before interpreting the correlation coefficient
  - Correlation is only applicable for linear relationships.
  - Correlation is **NOT** causation. A high correlation does not imply that changes in one variable cause changes in the other.
</div>

- **Formula:**
$$
r = \frac{\text{Cov}(X,Y)}{s_x s_y}
$$

| Correlation (r) | Strength        | Interpretation |
|------------------|-----------------|----------------|
| 0.00 – 0.10      | None            | No linear relationship |
| 0.10 – 0.30      | Weak            | Slight trend, not reliable |
| 0.30 – 0.50      | Moderate        | Noticeable relationship |
| 0.50 – 0.70      | Strong          | Clear relationship |
| 0.70 – 0.90      | Very strong     | Highly related variables |
| 0.90 – 1.00      | Near perfect    | Almost perfectly linear |
| -0.10 – 0.00     | None (negative) | No linear relationship |
| -0.30 – -0.10    | Weak negative   | Slight inverse trend |
| -0.50 – -0.30    | Moderate negative | Noticeable inverse relationship |
| -0.70 – -0.50    | Strong negative | Clear inverse relationship |
| -1.00 – -0.70    | Very strong negative | Highly inverse relationship |
| -1.00            | Perfect negative | Perfect inverse linear relationship |

<div class="container">

  <div class="box">

**Variable A vs B**
  ```{r} 
  cor(A, B)
  ```
  </div> 
 
</div>

<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
  <strong>Rule of thumb:</strong><br>

  if $|r| \geq \frac{2}{\sqrt(n)}$, the correlation is considered strong.
</div>

---
