# Appendix A: Bessel's Correction {#appendix-a}

```{r, echo=FALSE, include=FALSE}
library(ayuda)
library(ggplot2)
library(patchwork)

### Standard deviation
pop_sd <- function(vec, n = 0) {
  return(sqrt(sum((vec - mean(vec))^2) / (length(vec) - n))) # nolint
}

plotsd <- function(df, title_text) {
  df %>% ggplot(aes(x = iteration, y = sdd)) +
    geom_point(color = "blue", size = 2) +
    labs(x = "Iteration", y = "Standart deviation", title = title_text) +
    geom_hline(yintercept = std, linetype = "dashed",
               color = "red", linewidth = 1) +
    theme_gray() +
    theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))
}
```

Given a population of 50,000 random numbers between -1 and 1, we can calculate the standard deviation of the population. 
```{r}
set.seed(123)
pop <- runif(50000, -1, 1)
std <- pop_sd(pop)
```
Then, we can take multiple samples of size 50 from this population and calculate the standard deviation of each sample using different formulas: 
dividing by n, n-1, and n-2.
```{r} 
sds <- map(1:100, ~ mean(sapply(1:1000, function(x) pop_sd(sample(pop, 50), n=0)))) %>% unlist() # nolint
sds1 <- map(1:100, ~ mean(sapply(1:1000, function(x) pop_sd(sample(pop, 50), n=1)))) %>% unlist() # nolint
sds2 <- map(1:100, ~ mean(sapply(1:1000, function(x) pop_sd(sample(pop, 50), n=2)))) %>% unlist() # nolint
```
Finally, we can visualize the results to see how the choice of denominator affects the estimated standard deviation.
```{r}
df <- data.frame(
  iteration = 1:length(sds),
  sdd = sds
)
df1 <- data.frame(
  iteration = 1:length(sds1),
  sdd = sds1
)
df2 <- data.frame(
  iteration = 1:length(sds2),
  sdd = sds2
)

plotsd(df, "n") / plotsd(df1, "n-1") / plotsd(df2, "n-2")
```