# Simple Linear Regression

```{r, echo=FALSE, include=FALSE}
library(here)
library(ayuda)
library(ggplot2)
source(here::here("R/init.R"))
```

```{r, echo=FALSE}
#Dataset prep
adlb <- adam$adlb() %>% 
  filter(SAFFL == "Y" &  AVISITN==9999 & ANL01FL == "Y" & !(PARAMCD %in% c("COLOR"))) %>%  # nolint
  select(USUBJID, AGE, TRT01P, PARCAT1, PARAM, PARAMCD, PARAMN, AVISIT, AVISITN, AVAL, CHG, BASE) # nolint

UNIQ_PARAMs <- adlb  %>% select(PARCAT1, PARAM, PARAMCD, PARAMN)  %>% unique()

## List unique Parameters
adlb_wide <- adlb %>%
  select(USUBJID, AGE, TRT01P, PARAMCD, CHG, BASE, AVAL, AVISIT) %>%
  pivot_wider(
    id_cols = c(USUBJID, AGE, TRT01P, AVISIT),
    names_from = PARAMCD,
    values_from = c(AVAL, CHG, BASE)
  )

n <- length(adlb_wide$BASE_CA)
A <- adlb_wide$BASE_CA
B <- adlb_wide$CHG_CA
```

Simple linear regression is a statistical method that allows us to summarize and study relationships between two continuous (quantitative) variables. 
One variable, denoted $x$, is regarded as the predictor, explanatory, or independent variable. 
The other variable, denoted $y$, is regarded as the response, outcome, or dependent variable. 
The goal of simple linear regression is to find the best-fitting straight line through the points on a scatter plot of $x$ and $y$. 
This line is called the regression line. 

$$
y = \beta_0 + \beta_1 x + \varepsilon 
$$
where:

- $\beta_0$ intercet population parameter;
- $\beta_1$ slope population parameter;
- $\varepsilon$ error term.

The mean or expected value of $y$ for a given value of $x$.
$$
E(y|x) = \beta_0 + \beta_1 x
$$

<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
  Note $E(\varepsilon) = 0$ so the regression line captures the average relationship correctly.
</div>

We almost never have a population parameter to work with, so we need to estimate the parameters $b_0$ and $b_1$ using sample data.

$$
\hat{y} = b_0 + b_1 x 
$$

---

## Before regression: Exploratory Data Analysis

**Step 1** {#scatter_plot}
Plot the data to see if there is a relationship between the two variables. 

```{r, echo=FALSE}
# plot results by TRT01p as color
adlb_wide %>% 
  ggplot(aes(BASE_CA, CHG_CA, color = TRT01P)) +
  geom_point()
```

**Step 2**
Look for a visual relationship between the two variables. If there is a relationship, we can fit a line to the data.

```{r, echo=FALSE} 
adlb_wide %>% 
  ggplot(aes(BASE_CA, CHG_CA)) +
  geom_point()+
  geom_abline(slope = -0.6, intercept = 1.4, linetype = "dashed") +
  geom_abline(slope = -0.65, intercept = 1.5, linetype = "dotted")
```

**Step 3**
Calculate correlation between the two variables to see if there is a linear relationship between them.
For more information see \@ref(correlation).
```{r, echo=FALSE}
cor(adlb_wide$BASE_CA, adlb_wide$CHG_CA) 
```

**Step 4**
Calculate the centroid $\bar{x}$, $\bar{y}$

<div class="container">

  <div class="box">

**Variable A**
  ```{r}
  mean(adlb_wide$BASE_CA)  
  ```
  </div> 

  <div class="box">

**Variable B**
  ```{r}
  mean(adlb_wide$CHG_CA) 
  ```
  </div>
</div>

**Step 5**
Calculate the slope and intercept of the regression line using the formulas below.

---

## Least squares

We are looking for the line that best fits the data points. 
$$
\hat{y} = b_0 + b_1 x 
$$

In order to do that we are using the method of least squares, which minimizes the sum of the squared residuals (the vertical distances between the observed values and the predicted values).

$$
min \sum_{i=1}^{n} (y_i - \hat{y_i})^2
$$

where:

- $y_i$ observer values;
- $\hat{y_i} = b_0 + b_1 x_i$ estimated(predicted) values. 

or using residuals: 

$$
RSS = \sum_{i=1}^{n} e_i^2
$$

where $e_i = y_i - \hat{y_i}$ the $i$th residual this is the difference between the $i$th observed response value and the $i$th response value.

equivalently

$$
RSS = \sum_{i=1}^{n} (y_i - b_0 - b_1 x_i)^2
$$

The least squares approach chooses $b_0$ and $b_1$ to minimize the $RSS$. Using some calculus:

$$
\frac{\partial RSS}{\partial b_0} = -2 \sum_{i=1}^{n} (y_i - b_0 - b_1 x_i) = 0
$$

$$
\frac{\partial RSS}{\partial b_1} = -2 \sum_{i=1}^{n} x_i (y_i - b_0 - b_1 x_i) = 0
$$

Simplify:

$$
\sum y_i = n b_0 + b_1 \sum x_i \tag{1} 
$$

$$
\sum x_i y_i = b_0 \sum x_i + b_1 \sum x_i^2  \tag{2} 
$$

from (1):

$$
b_0 = \frac{\sum y_i - b_1 \sum x_i}{n}
$$

Substitute $b_0$ into (2):

$$
\sum x_i y_i = \left( \frac{\sum y_i - b_1 \sum x_i}{n} \right) \sum x_i + b_1 \sum x_i^2
$$

Expand:

$$
\sum x_i y_i = \frac{\sum x_i \sum y_i}{n} - b_1 \frac{(\sum x_i)^2}{n} + b_1 \sum x_i^2
$$

Rearrange terms:

$$
\sum x_i y_i - \frac{\sum x_i \sum y_i}{n} = b_1 \left( \sum x_i^2 - \frac{(\sum x_i)^2}{n} \right)
$$

Solve for $b_1$:

$$
b_1 = \frac{ \sum x_i y_i - \frac{\sum x_i \sum y_i}{n} }{ \sum x_i^2 - \frac{(\sum x_i)^2}{n}}
$$

Equivalent centered form:

$$
b_1 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^{n}(x_i - \bar{x})^2} \tag{3} 
$$

and from (1)

$$
b_0 = \bar{y} - b_1 \bar{x} \tag{4} 
$$

<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
  <strong>Cozy correlation:</strong><br>

  $$ b_1 = r \frac{s_y}{s_x} $$

  where:

  - $s_x$ standard deviation of $x$;
  - $s_y$ standard deviation of $y$;
</div>

---

## Regression line

$$
\hat{y} = b_0 + b_1 x 
$$

Find $b_0$ and $b_1$ using (3) and (4):

```{r , echo=FALSE} 
fit <- adlb_wide %>% lm(CHG_CA ~ BASE_CA, data = .)
x <- adlb_wide$BASE_CA
y <- adlb_wide$CHG_CA
yhat <- fitted(fit)
resid <- residuals(fit)
ybar <- mean(y)
```

```{r results='asis'} 
b_1  <- coef(fit)[2]
b_0 <- coef(fit)[1]
sprintf("$$b_0 = %.2f$$", round(b_0, 2)) %>% cat()
sprintf("$$b_1 = %.2f$$", round(b_1, 2)) %>% cat()
```

```{r, echo=FALSE}
label_eq <- sprintf(
  "$$\\hat{y_i} = %.2f + %.2f x_i$$",
  round(b_0, 2),
  round(b_1, 2)
)
```
```{r results='asis'}
label_eq %>% cat()
```

```{r, echo=FALSE}
adlb_wide %>% 
  ggplot(aes(BASE_CA, CHG_CA)) +
  geom_point(aes(color = "Observed data")) +
  geom_abline(aes(slope = b_1, intercept = b_0,
                  color = "Regression line")) +
  geom_point(
    data = adlb_wide %>% summarise(
      BASE_CA = mean(BASE_CA, na.rm = TRUE),
      CHG_CA  = mean(CHG_CA, na.rm = TRUE)
    ), aes(BASE_CA, CHG_CA, color = "Centroid")) +
  # legend colors
  scale_color_manual( name = "Components",
    values = c(
      "Observed data" = "black",
      "Regression line" = "red",
      "Centroid" = "yellow"
    )
  )
```

---

## Understanding model error

In simple linear regression, the total variation in the response variable is decomposed into three components:


### Total Sum of Squares (SST)

Measures the total variability in the observed data:

$$
SST = \sum_{i=1}^{n}(y_i - \bar{y})^2
$$

```{r results='asis'} 
SST <- sum((y - ybar)^2)
sprintf("$$SST = %.2f$$", round(SST, 2)) %>% cat() 
```

### Regression Sum of Squares (SSR)

Measures the explained variation by the regression model:

$$
SSR = \sum_{i=1}^{n}(\hat{y_i} - \bar{y})^2
$$

```{r results='asis'} 
SSR <- sum((yhat - ybar)^2)
sprintf("$$SSR = %.2f$$", round(SSR, 2)) %>% cat() 
```

### Error Sum of Squares (SSE)

Measures the unexplained variation (residual error):

$$
SSE = \sum_{i=1}^{n}(y_i - \hat{y_i})^2
$$

```{r results='asis'} 
SSE <- sum(resid^2)
sprintf("$$SSE = %.2f$$", round(SSE, 2)) %>% cat() 
```

### Key Identity

The total variation is decomposed as:

$$
SST = SSR + SSE
$$

```{r}
SSR + SSE 
```

### Coefficient of Determination ($R^2$)

Another way to measure the contribution of $x$ in predicting $y$ is to consider how much the errors of prediction can be reduced by using the information provided by $x$.

$$
R^2 = \frac{SSR}{SST} = \frac{SST - SSE}{SST} = 1 - \frac{SSE}{SST} = 1 - \frac{\sum (y_i - \hat{y_i})^2}{\sum (y_i - \bar{y})^2} 
$$

```{r results='asis'}
sprintf("$$R^2 = %.3f$$", round(SSR/SST, 3)) %>% cat()
summary(fit)$r.squared 
```

<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
  <strong> Practical interpratation of the Coefficient of Determination:</strong><br>
  About $ 100(R^2) % $ for the total sum of squeres of deviations of the sample $y$ values about their mean $\bar{y}$ can be explained by using $x$ to predict 
  $y$ in the straight line regression model. The remaining $100(1-R^2) %$ of the total sum of squares is due to the variability in $y$ that cannot be explained by the linear relationship with $x$. 
</div>

### Adjusted R-squared

Adjusted \(R^2\) accounts for the number of predictors in the model and penalizes unnecessary complexity.

$$
R^2_{adj} = 1 - \frac{(1 - R^2)(n - 1)}{n - p - 1}
$$

where:

- $p$  = number of predictors. Sometimes instead of $n - p - 1$ you may see $df$
- $df = n - p - 1$ is degrees of freedom

```{r results='asis'}
p <- 1 # number of predictors 
df <- n - p - 1
R2_adj <- 1 - ((1 - SSR/SST) * (n - 1)) / (n - p - 1)
sprintf("$$R^2_{adj} = %.3f$$", round(R2_adj, 3)) %>% cat()
summary(fit)$adj.r.squared
```

---

## Standardized Variables

Let us standardize the variables $x$ and $y$ so that they each have mean $0$ and standard deviation $1$.

The standardized versions are:

$$
z_{x,i} = \frac{x_i - \bar{x}}{s_x}
$$

and

$$
z_{y,i} = \frac{y_i - \bar{y}}{s_y}
$$

where:

- $\bar{x}$ and $\bar{y}$ are the sample means
- $s_x$ and $s_y$ are the sample standard deviations

After standardization:

$$
E(z_x) = 0
\qquad
E(z_y) = 0
$$

and

$$
SD(z_x) = 1
\qquad
SD(z_y) = 1
$$

```{r echo=FALSE}
# Standardize variables
adlb_std <- adlb_wide %>%
  mutate(
    z_BASE_CA = scale(BASE_CA)[,1],
    z_CHG_CA  = scale(CHG_CA)[,1]
  )

# Plot standardized variables
adlb_std %>%
  ggplot(aes(z_BASE_CA, z_CHG_CA )) +
  geom_point(size = 2, alpha = 0.8) +
  labs(
    title = "Standardized Variables",
    x = expression(z[x]),
    y = expression(z[y]) 
  ) 
```

### Correlation

Lets check the correlation between the standardized variables:

```{r results='asis'}
cor(adlb_wide$BASE_CA, adlb_wide$CHG_CA)  %>% round(3) %>% 
  sprintf("Original correlation $$r = %.3f$$", .) %>% cat()
cor(adlb_std$z_BASE_CA, adlb_std$z_CHG_CA)  %>% round(3) %>%
  sprintf("Correlation of standardized variables $$r = %.3f$$", .) %>% cat()
```

We can see that correlatio between the original variables and the standardized variables is the same.

### Standardized Regression

Consider the simple linear regression model after standardizing both variables the standardized regression model becomes:

$$
z_{y,i} = b_1^{*} z_{x,i} + \varepsilon_i
$$

where:

$$
b_1^{*} = r_{xy}
$$

is the sample correlation between $x$ and $y$.
Since standardized variables have mean $0$:

$$
b_0^{*} = 0
$$

Thus, the standardized regression line passes through the origin.


```{r, echo=FALSE}
fit_z <- adlb_std %>% lm(z_CHG_CA ~ z_BASE_CA, data = .)

adlb_std %>%
  ggplot(aes(z_BASE_CA, z_CHG_CA )) +
  geom_point(aes(color = "Observed data"), size = 2, alpha = 0.8)  +
  geom_abline(aes(slope = coef(fit_z)[2], intercept = coef(fit_z)[1],
                  color = "Standardized Regression line")) + 
  scale_color_manual( name = "Components",
    values = c(
      "Observed data" = "black",
      "Standardized Regression line" = "red" 
    )
  )+ labs(
    title = "Standardized Variables",
    x = expression(z[x]),
    y = expression(z[y]) 
  ) 
```
<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
  <strong> Using standardized scores can help:</strong><br>
  
  - interprete variables with different scales;
  - helpful to see general patterns in the data;
  - help to see which variable offer more to keep and which to drop;
  - determine whiich variable in multiple regression account for more variance in the response variable.
</div>

---

## Statistical sagnificance

* Overal $F$-test or $t$ test is sagnificant;
* Can we reject $H_0$ that the slope $b_1$ is zero;
* Does confidence inteval for $b_1$ contain zero.

### Standart error of estimate

$$
MSE = \frac{SSE}{n-2}
$$

where:

- $SSE$ is the sum of squared errors
- $MSE$ is the mean squared error

```{r results='asis'}
MSE <- SSE/(n-2)
sprintf("$$MSE = %.3f$$", round(MSE, 3)) %>% cat()
summary(fit)$sigma^2
```

### Confidence interval for slope

For the simple linear regression model

the $(1 - \alpha)100\%$ confidence interval for the slope $b_1$ is

$$
b_1 \pm t_{\alpha/2,\,df} \cdot SE(b_1)
$$

where:

- $b_1$ is the estimated slope
- $SE(b_1)$ is the standard error of the slope estimate
- $t_{\alpha/2,\,df}$ is the critical value from the $t$ distribution
- $\alpha$ usually is set to $0.05$ for a $95\%$ confidence interval
- $df$ are the degrees of freedom

The standard error of the slope is

$$
SE(b_1) =
\sqrt{
\frac{MSE}
{\sum (x_i - \bar{x})^2}
}
$$

```{r results='asis'}
SE_b1 <- sqrt(MSE/((n-1)*var(x, na.rm = TRUE)))
sprintf("$$SE(b_1) = %.5f$$", round(SE_b1, 6)) %>% cat()
summary(fit)$coefficients[2, 2]
```

Confidence interval for the slope:

```{r results='asis'}
alpha <- 0.05

t_crit <- qt(1 - alpha/2, df)

lower <- b_1 - t_crit * SE_b1
upper <- b_1 + t_crit * SE_b1

cat(sprintf("$$[%.3f, %.3f]$$", lower, upper))

confint(fit, level = 0.95) [2,]
```

The confidence interval does not contain zero.

### Test statistic

#### t-statistic (for a coefficient)

For two-tailed testing $H_0: b_1 = 0$:

$$
t = \frac{b_1}{SE(b_1)}
$$

where:

- $b_1$ is the estimated coefficient  
- $SE(b_1)$ is its standard error  


at significance level $\alpha$, we reject $H_0$ if:

$$
|t| > t_{\alpha/2,\, df}
$$

```{r results='asis'}
t  <- b_1/SE_b1
cat(sprintf("$$ |%.3f| > %.3f$$", abs(t), t_crit))


t_val <- summary(fit)$coefficients[2, "t value"]
p_val <- summary(fit)$coefficients[2, "Pr(>|t|)"]

cat(sprintf("$$t = %.3f, \\quad p = %.6f$$", t_val, p_val))
```

The confidence interval does not contain zero and $|t| > t_{\alpha/2,\, df}$ , so we can reject the null hypothesis that the slope is zero and conclude 
that there is a statistically significant linear relationship between $x$ and $y$.


#### F-statistic (overall regression significance)

For two-tailed testing:
$$
H_0: b_1 = 0
$$

$$
F = \frac{MSR}{MSE}
$$

where:

- $MSR = \frac{SSR}{p}$ (Mean Square Regression)
- $MSE = \frac{SSE}{n-p-1}$ (Mean Square Error)

Equivalent form:

$$
F = \frac{SSR / p}{SSE / (n - p - 1)}
$$

at significance level $\alpha$, we reject $H_0$ if:

$$
|F| > F_{\alpha,\, p,\, df}
$$

```{r results='asis'}
MSR <- SSR/p
F  <- MSR/MSE
F_crit  <-  qf(1 - alpha/2, p, df)

cat(sprintf("$$ |%.3f| > %.3f$$", abs(F), F_crit))

summary(fit)$fstatistic[1]
```

The confidence interval does not contain zero and $|F| > F_{\alpha,\, p,\, df}$ , so we can reject the null hypothesis that the slope is zero and conclude 
that there is a statistically significant linear relationship between $x$ and $y$.

---

## Bands

### Confidence band
<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
**NOTE**: Regression is not deterministic.
</div>
We can not generate an exact value but we can generate mean and confidence interval for 
the mean response at a given predictor value $x_0$.
For a given predictor value $x_0$, the confidence interval for the mean response is

$$
\hat{y}(x_0)
\pm
t_{\alpha/2,\,n-2}
\sqrt{
MSE
\left(
\frac{1}{n}
+
\frac{(x_0 - \bar{x})^2}
{\sum (x_i - \bar{x})^2}
\right)
}
$$

where:

- $\hat{y}(x_0)$ is the fitted mean response at $x_0$
- $MSE = \dfrac{SSE}{n-2}$
- $t_{\alpha/2,\,df}$ is the critical $t$ value
- $n$ is the number of observations

From the formula we can see that the most precice estimate of the population mean is at 
the centroid of the data ($x_0 = \bar{x}$). After the $\bar{x}$ the interval become vider.

```{r results='asis'}
xbar <- mean(x, na.rm = TRUE)
x0 <- sample(x, 1) # take a random value 
y_x0 <- predict( fit, newdata = data.frame( BASE_CA = x0)) # nolint


SE_x0 <- sqrt(MSE * ((1 / n) + ((x0 - xbar)^2) / sum((x - xbar)^2)))

lower <- y_x0 - t_crit * SE_x0
upper <- y_x0 + t_crit * SE_x0

cat(sprintf("$$[%.3f, %.3f]$$", lower, upper))

predict(
  fit,
  newdata = data.frame(BASE_CA = x0),
  interval = "confidence",
  level = 0.95
)
```

```{r}
adlb_wide %>% 
  ggplot(aes(BASE_CA, CHG_CA)) +
  
  geom_point(aes(color = "Observed data")) +
  
  geom_smooth(
    method = "lm",
    se = TRUE,
    aes(color = "Regression line",
        fill = "95% Confidence Band")
  ) +
  
  geom_point(
    data = adlb_wide %>% summarise(
      BASE_CA = mean(BASE_CA, na.rm = TRUE),
      CHG_CA  = mean(CHG_CA, na.rm = TRUE)
    ),
    aes(BASE_CA, CHG_CA, color = "Centroid"),
    size = 2
  ) +
  
  scale_color_manual(
    name = "Components",
    values = c(
      "Observed data" = "black",
      "Regression line" = "red",
      "Centroid" = "yellow"
    )
  ) +
  
  scale_fill_manual(
    name = "Components",
    values = c(
      "95% Confidence Band" = "skyblue"
    )
  )
```

### Prediction band

For a given predictor value $x_0$, the prediction interval for a new observation is

$$
\hat{y}(x_0)
\pm
t_{\alpha/2,\,df}
\sqrt{
MSE
\left(
1 +
\frac{1}{n}
+
\frac{(x_0 - \bar{x})^2}
{\sum (x_i - \bar{x})^2}
\right)
}
$$

Important Difference
<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
Important Difference
Prediction bands are wider than confidence bands because they include:

- uncertainty in estimating the mean response
- variability of individual observations
</div>


```{r results='asis'}
PSE_x0 <- sqrt(MSE + SE_x0^2)

lower <- y_x0 - t_crit * PSE_x0
upper <- y_x0 + t_crit * PSE_x0

cat(sprintf("$$[%.3f, %.3f]$$", lower, upper))

predict(
  fit,
  newdata = data.frame(BASE_CA = x0),
  interval = "prediction",
  level = 0.95
)
```

```{r}

grid <- data.frame(
  BASE_CA = seq(
    min(adlb_wide$BASE_CA, na.rm = TRUE),
    max(adlb_wide$BASE_CA, na.rm = TRUE),
    length.out = 100
  )
)

# confidence + prediction
ci <- predict(fit, newdata = grid, interval = "confidence")
pi <- predict(fit, newdata = grid, interval = "prediction")

grid_ci <- cbind(grid, ci)
grid_pi <- cbind(grid, pi)
 
ggplot() +
  
  # prediction band (must come first)
  geom_ribbon(
    data = grid_pi,
    aes(x = BASE_CA, ymin = lwr, ymax = upr, fill = "95% Prediction Interval"),
    alpha = 0.15
  ) +
  
  # confidence band
  geom_ribbon(
    data = grid_ci,
    aes(x = BASE_CA, ymin = lwr, ymax = upr, fill = "95% Confidence Band"),
    alpha = 0.25
  ) +
  
  # raw data
  geom_point(
    data = adlb_wide,
    aes(BASE_CA, CHG_CA, color = "Observed data")
  ) +
  
  # regression line
  geom_smooth(
    data = adlb_wide,
    aes(BASE_CA, CHG_CA, color = "Regression line"),
    method = "lm",
    se = FALSE
  ) +
  
  scale_color_manual(
    name = "Components",
    values = c(
      "Observed data" = "black",
      "Regression line" = "red"
    )
  ) +
  
  scale_fill_manual(
    name = "Intervals",
    values = c(
      "95% Confidence Band" = "skyblue",
      "95% Prediction Interval" = "orange"
    )
  )
```

---

## Residual analysis

Residual analysis is used to assess whether the assumptions of the regression model are reasonable.
The variance left unexplained is due to the error term $\varepsilon$ in the regression model.

The residual for observation $i$ is

$$
e_i = y_i - \hat{y}_i
$$

where:

- $y_i$ is the observed response
- $\hat{y_i}$ is the fitted value

```{r}
ggplot(
  data.frame(
    fitted = fitted(fit),
    residuals = residuals(fit)
  ),
  aes(fitted, residuals)
) +
  geom_point() +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residuals vs Fitted",
    x = "Fitted values",
    y = "Residuals"
  ) 
```

### Homoscedasticity

Homoscedasticity means that the variance of the errors is constant:

$$
Var(\varepsilon_i) = \sigma^2
$$

for all observations.

A homoscedastic residual plot typically looks like a horizontal cloud of points.
```{r, echo=FALSE}
set.seed(123)

N <- 200
X <- runif(N, 0, 10)

Y <- 5 + 2 * X + rnorm(N, mean = 0, sd = 1)

test_fit <- lm(Y ~ X)

# Residual plot
ggplot(
  data.frame(
    fitted = fitted(test_fit),
    residuals = residuals(test_fit)
  ),
  aes(fitted, residuals)
) +
  geom_point() +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             color = "red") +
  labs(
    title = "Homoscedastic Residual Plot",
    x = "Fitted values",
    y = "Residuals"
  ) 
```


### Heteroscedasticity

Heteroscedasticity occurs when the variance of the errors changes with the fitted values or predictors.

In a residual plot:

- spread increases or decreases
- funnel or cone shapes may appear
- residual variability is not constant

Common patterns:

- widening spread
- narrowing spread
- curved structures

```{r, echo=FALSE}
# Error variance increases with x
Y <- 5 + 2 * X + rnorm(N, mean = 0, sd = 0.5 * X)

test_fit <- lm(Y ~ X)

# Residual plot
ggplot(
  data.frame(
    fitted = fitted(test_fit),
    residuals = residuals(test_fit)
  ),
  aes(fitted, residuals)
) +
  geom_point() +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             color = "red") +
  labs(
    title = "Heteroscedastic Residual Plot",
    x = "Fitted values",
    y = "Residuals"
  )
```


### Desired Residual Plot

A good residual plot should show:

- random scatter around $0$
- no systematic structure
- approximately equal variance
- no strong outliers

### Tests for residuals

#### Shapiro-Wilk Test

The Shapiro-Wilk test is used to assess whether residuals are normally distributed.

**Hypotheses**

$$
H_0:\ \text{Residuals are normally distributed}
$$

$$
H_A:\ \text{Residuals are not normally distributed}
$$

```{r}
# Shapiro-Wilk test
shapiro.test(residuals(fit))
```

**Interpretation**

- Large p-value:
  residuals appear approximately normal

- Small p-value:
  evidence against normality


#### Breusch-Pagan Test

The Breusch-Pagan test is used to detect heteroscedasticity.

**Hypotheses**

$$
H_0:\ \text{Constant variance (homoscedasticity)}
$$

$$
H_A:\ \text{Non-constant variance (heteroscedasticity)}
$$

```{r}
# Breusch-Pagan test 
lmtest::bptest(fit)
```

**Interpretation**

- Large p-value:
  constant variance is reasonable

- Small p-value:
  evidence of heteroscedasticity

#### Durbin-Watson Test

The Durbin-Watson test is used to detect autocorrelation in residuals.

**Hypotheses**

$$
H_0:\ \text{No autocorrelation}
$$

$$
H_A:\ \text{Autocorrelation exists}
$$

```{r}
# Durbin-Watson test
lmtest::dwtest(fit)
```

**Interpretation**

- Large p-value:
  residuals appear independent

- Small p-value:
  evidence of autocorrelation

#### White Test

The White test is a general test for heteroscedasticity.

It allows variance to depend on:
- predictors
- squared predictors
- interaction terms

**Hypotheses**

$$
H_0:\ \text{Homoscedasticity}
$$

$$
H_A:\ \text{Heteroscedasticity}
$$

```{r}
# White test
lmtest::bptest(
  fit,
  ~ fitted(fit) + I(fitted(fit)^2)
)
```

**Interpretation**

- Large p-value:
  no evidence of heteroscedasticity

- Small p-value:
  evidence of heteroscedasticity

#### NCV Test (Non-Constant Variance Test)

The NCV test evaluates whether residual variance changes with fitted values.

**Hypotheses**

$$
H_0:\ \text{Constant variance}
$$

$$
H_A:\ \text{Non-constant variance}
$$

```{r}
# NCV test
car::ncvTest(fit)
```

**Interpretation**

- Large p-value:
  constant variance assumption is reasonable

- Small p-value:
  evidence of changing variance




---


## Outliers

- **Always do scatter plot!!!** Sometimes you will see points out of place. \@ref(scatter_plot) 
- Also chech other methods to identify outliers, see \@ref(iqr).
- An out of normal point may have a large residual.

<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;"> 
**NOTE**: for standardised residuals usually $<2$ are more possible otliers.
</div>

```{r}
ggplot(
  data.frame(
    fitted = fitted(fit_z),
    residuals = residuals(fit_z)
  ),
  aes(fitted, residuals)
) +
  geom_point() +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Standardised Residuals vs Fitted",
    x = "Fitted values",
    y = "Residuals"
  ) 
```

### Influential observations

Influential observations are data points that have a strong effect on the fitted regression model.

Outliers can influence the regression line and pull it toward them, potentially changing:

- slope estimates
- intercept estimates
- hypothesis tests
- predictions

Not all outliers are influential.

### Leverage Points

Leverage measures how far a predictor value is from the mean of the predictor.

Observations with unusual predictor values may have high leverage and can strongly affect the regression line.

For simple linear regression, the leverage of observation $i$ is

$$
h_i =
\frac{1}{n}
+
\frac{(x_i - \bar{x})^2}
{\sum_{j=1}^{n}(x_j - \bar{x})^2}
$$

where:

- $n$ is the sample size
- $\bar{x}$ is the sample mean of $x$
- $h_i$ is the leverage of observation $i$

 Rule of Thumb

<div style="border-left: 4px solid #2c7be5; background:#f5f9ff; padding:12px; border-radius:6px;">
  <strong> Rule of Thumb</strong><br>
  Potential high leverage points are often identified using:
  $$
  h_i > \frac{2(p+1)}{n}
  $$

  or

  $$
  h_i > \frac{3(p+1)}{n}
  $$

  where:

  - $p$ is the number of predictors
  - $n$ is the sample size

</div>
 
```{r}
lev <- hatvalues(fit)

# Add leverage to dataset
adlb_wide$leverage <- lev

cutoff <- 2 * ((p+1) / nrow(adlb_wide))

# Show highest leverage points
ggplot(adlb_wide,
       aes(BASE_CA, leverage,
           color = leverage > cutoff)) +
  geom_point(size = 2) +
  geom_hline(
    yintercept = cutoff,
    color = "red",
    linetype = "dashed"
  ) +
  labs(
    title = "High Leverage Points",
    x = "BASE_CA",
    y = "Leverage"
  )
```